\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-08-23 paper.qmd
Added notes to what to write in Section 6

The \gD{} layouts generated by tSNE, UMAP, and PaCMAP show five well-separated clusters which evident these methods effectively preserve the global structure. In tSNE (@fig-gau-tsne-sc (a)), these clusters appear closely. UMAP arranges all clusters in a parallel manner, with three aligned in one line and the other two in a separate line (@fig-gau-umap-sc (a)). In contrast, PaCMAP shows one central cluster and the remaining four spread out in different directions (@fig-gau-pacmap-sc (a)).

The tSNE and UMAP shows *filled out* clusters which provide evidence that these methods preserve the local structure (@fig-gau-tsne-sc (c) and @fig-gau-umap-sc (c)). On the other hand, PaCMAP shows *flat* shapes clusters in the model and evident that PaCMAP fail to capture the within-cluster variation (@fig-gau-pacmap-sc (c)).

@@

## Compute hexbin parameters
num_bins_x_gau1 <- 14
lim1 <- gau1_scaled_obj$lim1
lim2 <- gau1_scaled_obj$lim2
r2_gau1 <- diff(lim2)/diff(lim1)

gau1_model <- fit_highd_model(
  training_data = training_data_gau,
  emb_df = tsne_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x",
  q = 0.16
)

@@

## Hexagonal binning to have regular hexagons
hb_obj_gau1 <- hex_binning(
  data = tsne_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  q = 0.16)

@@

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_gau1$hexID, df_b$hb_id),] |>
  dplyr::select(-hb_id)

@@

points_only_df <- df_exe |>
    dplyr::select(-type)

projected <- as.matrix(points_only_df) %*% projection

projected_df <- projected |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  dplyr::mutate(type = df_exe$type) |>
  dplyr::mutate(ID = dplyr::row_number())

proj_model <- projected_df |>
  dplyr::filter(type == "model")

model_df <- dplyr::left_join(
  distance_df_small_edges |> select(-distance), 
  proj_model, 
  by = c("from" = "ID"))

names(model_df)[3:NCOL(model_df)] <- paste0(names(proj_model)[-NCOL(proj_model)], "_from")

model_df <- dplyr::left_join(model_df, proj_model, by = c("to" = "ID"))
names(model_df)[(2 + NCOL(proj_model)):NCOL(model_df)] <- paste0(names(proj_model)[-NCOL(proj_model)], "_to")

five_gau_proj_tsne_model <- projected_df |>
  filter(type == "data") |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    alpha = 0.3) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  xlim(c(-0.18, 0.1)) +
  ylim(c(-0.78, -0.55)) +
  interior_annotation("a", 
                      position = c(0.08, 0.9),
                      cex = 3)

@@

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_gau1$hexID, df_b$hb_id),] |>
  dplyr::select(-hb_id)

@@

df_b <- df_b |>
  select(-type)

@@

gau1_model <- fit_highd_model(
  training_data = training_data_gau,
  emb_df = pacmap_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x",
  q = 0.14
)

@@

## Hexagonal binning to have regular hexagons
hb_obj_gau1 <- hex_binning(
  data = pacmap_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  q = 0.14)

@@

points_only_df <- df_exe |>
    dplyr::select(-type)

projected <- as.matrix(points_only_df) %*% projection

projected_df <- projected |>
  tibble::as_tibble(.name_repair = "unique") |>
  dplyr::rename(c("proj1" = "...1",
                  "proj2" = "...2")) |>
  dplyr::mutate(type = df_exe$type) |>
  dplyr::mutate(ID = dplyr::row_number())

proj_model <- projected_df |>
  dplyr::filter(type == "model")

model_df <- dplyr::left_join(
  distance_df_small_edges |> select(-distance), 
  proj_model, 
  by = c("from" = "ID"))

names(model_df)[3:NCOL(model_df)] <- paste0(names(proj_model)[-NCOL(proj_model)], "_from")

model_df <- dplyr::left_join(model_df, proj_model, by = c("to" = "ID"))
names(model_df)[(2 + NCOL(proj_model)):NCOL(model_df)] <- paste0(names(proj_model)[-NCOL(proj_model)], "_to")

five_gau_proj_pacmap_model <- projected_df |>
  filter(type == "data") |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    alpha = 0.3) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  xlim(c(-0.18, 0.1)) +
  ylim(c(-0.78, -0.55)) +
  interior_annotation("c", 
                      position = c(0.08, 0.9),
                      cex = 3)

>>>>

XXX Add tSNE, UMAP, and PaCMAP NLDR layouts and describe what is happening
XXX Discuss the models of one specific cluster (filled out, pancakes): tSNE, UMAP, PaCMAP are useful to capture global structures, but this is an evidence where PaCMAP fail to capture the local structure with default hyper-parameter settings than tSNE and UMAP
XXX Why? hyper-parameter choice specially number of neighbors, data characteristics (intrinsic dimensionality), imbalance between global and local structures, optimization and initialization issues

<!-- The \gD{} layouts generated by tSNE, UMAP, and PaCMAP show five well-separated clusters which evident these methods effectively preserve the global structure. In tSNE (@fig-gau-tsne-sc (a)), these clusters appear closely. UMAP arranges all clusters in a parallel manner, with three aligned in one line and the other two in a separate line (@fig-gau-umap-sc (a)). In contrast, PaCMAP shows one central cluster and the remaining four spread out in different directions (@fig-gau-pacmap-sc (a)). -->

<!-- The tSNE and UMAP shows *filled out* clusters which provide evidence that these methods preserve the local structure (@fig-gau-tsne-sc (c) and @fig-gau-umap-sc (c)). On the other hand, PaCMAP shows *flat* shapes clusters in the model and evident that PaCMAP fail to capture the within-cluster variation (@fig-gau-pacmap-sc (c)).  -->

@@

## Compute hexbin parameters
num_bins_x_gau1 <- 13
lim1 <- gau1_scaled_obj$lim1
lim2 <- gau1_scaled_obj$lim2
r2_gau1 <- diff(lim2)/diff(lim1)

gau1_model <- fit_highd_model(
  training_data = training_data_gau,
  emb_df = tsne_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x",
  q = 0.1
)

@@

## Hexagonal binning to have regular hexagons
hb_obj_gau1 <- hex_binning(
  data = tsne_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  q = 0.1)

@@

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_gau1$hexID, df_b$hb_id),] |>
  dplyr::select(-hb_id) |>
  select(-type)

@@

five_gau_proj_tsne_model <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    alpha = 0.3) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.18, 0.1)) +
  ylim(c(-0.78, -0.55)) +
  interior_annotation("a", 
                      position = c(0.08, 0.9),
                      cex = 3)

@@

## Reorder the rows of df_b according to the hexID order in df_b_with_center_data
df_b <- df_b[match(df_bin_centroids_gau1$hexID, df_b$hb_id),] |>
  dplyr::select(-hb_id)|>
  select(-type)

@@

@@

gau1_model <- fit_highd_model(
  training_data = training_data_gau,
  emb_df = pacmap_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  is_bin_centroid = TRUE,
  is_rm_lwd_hex = FALSE,
  col_start_highd = "x",
  q = 0.1
)

@@

## Hexagonal binning to have regular hexagons
hb_obj_gau1 <- hex_binning(
  data = pacmap_gau_scaled,
  bin1 = num_bins_x_gau1,
  r2 = r2_gau1,
  q = 0.1)

@@

five_gau_proj_pacmap_model <- projected_df |>
  ggplot(
    aes(
      x = proj1, 
      y = proj2)) +
  geom_segment(
    data = model_df, 
    aes(
      x = proj1_from, 
      y = proj2_from, 
      xend = proj1_to, 
      yend = proj2_to), 
    color = "#33a02c") +
  geom_point(
    alpha = 0.3) +
  scale_color_manual(values = c("#000000", "#33a02c")) +
  geom_segment(
    data=axes, 
    aes(x=x1, y=y1, xend=x2, yend=y2), 
    colour="grey70") +
  geom_text(
    data=axes, 
    aes(x=x2, y=y2, label=rownames(axes)), 
    colour="grey50",
    size = 3) +
  geom_path(
    data=circle, 
    aes(x=c1, y=c2), colour="grey70") +
  coord_fixed() +
  xlim(c(-0.18, 0.1)) +
  ylim(c(-0.78, -0.55)) +
  interior_annotation("c", 
                      position = c(0.08, 0.9),
                      cex = 3)

### 2024-08-23 paper.tex
Regenerated Figure 2
Updated script to get the projections for UMAP
Add function to generate axes
Added notes to what to write in Section 6

The \gD{} layouts generated by tSNE, UMAP, and PaCMAP show five
well-separated clusters which evident these methods effectively preserve
the global structure. In tSNE (\textbf{?@fig-gau-tsne-sc} (a)), these
clusters appear closely. UMAP arranges all clusters in a parallel
manner, with three aligned in one line and the other two in a separate
line (\textbf{?@fig-gau-umap-sc} (a)). In contrast, PaCMAP shows one
central cluster and the remaining four spread out in different
directions (\textbf{?@fig-gau-pacmap-sc} (a)).

The tSNE and UMAP shows \emph{filled out} clusters which provide
evidence that these methods preserve the local structure
(\textbf{?@fig-gau-tsne-sc} (c) and \textbf{?@fig-gau-umap-sc} (c)). On
the other hand, PaCMAP shows \emph{flat} shapes clusters in the model
and evident that PaCMAP fail to capture the within-cluster variation
(\textbf{?@fig-gau-pacmap-sc} (c)).

>>>>

XXX Add tSNE, UMAP, and PaCMAP NLDR layouts and describe what is
happening XXX Discuss the models of one specific cluster (filled out,
pancakes): tSNE, UMAP, PaCMAP are useful to capture global structures,
but this is an evidence where PaCMAP fail to capture the local structure
with default hyper-parameter settings than tSNE and UMAP XXX Why?
hyper-parameter choice specially number of neighbors, data
characteristics (intrinsic dimensionality), imbalance between global and
local structures, optimization and initialization issues
